---
title: "class05_dataVisGgplot2"
author: "Derek Zhang (PID: A17819201)"
format: pdf
toc: true
---

## Background
there are many graphics systems in R for making plots
These include so-called "base" R graphics like the `plot()` function and add on packages like ggplot2.

We can use the in-built `cars` dataset:
```{r}
head(cars)
```

```{r}
plot(cars)
```

The main function in **ggplot2** packet is the ggplot, and we before we run it we must install ggplot2 by running 'install.packages("ggplot2")' in the console

> **N.B.** We never run `install.packages()` in our quarto doc because that would run it each time we render and become problematic

Once installed we need to load up the package into our R session
```{r}
#install.packages("ggplot2")
library(ggplot2)
ggplot(cars)
```

Every ggplot has at least 3 layers:
- The **data** (a data.frame of the stuff we want to plot)
- The **aesthetics** (how the data maps to the plot)
- The **geom** layer (how you want the plot drawn, e.g. points, lines, etc.)

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```

## Add some custom features

Lets add a trend line that shows the relationship between speed and distance.
```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() + 
  geom_smooth(method = "lm") +
  theme_light() +
  labs(title="Stopping Dist of Old Cars")
```

## Gene expression figure

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
sum(genes$State == "up")
genes[genes$State=="up",]
```
A useful new function in this context is the `table()` function:
```{r}
table(genes$State)
```
```{r}
ggplot(genes) +
  aes(x=Condition1, y=Condition2, color = State) +
  labs(title = "Gene Expression Changes Upon Drug Treatment", x = "Control (no drug)", y = "Drug Treatment", color = "Legend") +
  scale_colour_manual( values=c("blue","gray","red") ) +
  geom_point()
```

## Going further

Here we read the famous gapmider dataset
```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

> Q. How many entries (i.e. rows) are there in this dataset?
```{r}
nrow(gapminder)
```

> Q. How many countries are in this dataset
```{r}
length(table(gapminder$country))
```

```{r}
length(unique(gapminder$country))
```

Let's make our first plot of the dataset:
```{r}
p <- ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp) +
  geom_point(aes(color=continent), alpha = 0.3)
```

```{r}
p
```

```{r}
p + facet_wrap(~continent)
```

Now we will make a plot for years 1977 and 2007

```{r}
library(dplyr)
gapminder_1977_2007 <- gapminder %>% filter (year == 2007 | year == 1977)
ggplot(gapminder_1977_2007) +
  aes(x = gdpPercap, y = lifeExp) +
  geom_point(aes(color=continent), alpha = 0.3) +
  facet_wrap(~year)
```

> Q. Make a histogram of lifeExp faceted by continent and colored by continent

Faceted by continent:
```{r}
ggplot(gapminder) +
  aes(lifeExp) +
  geom_histogram() +
  facet_wrap(~continent)
```

Coloured by continent:
```{r}
ggplot(gapminder) +
  aes(lifeExp) +
  geom_histogram(aes(fill=continent))
```
## Optional: Bar Charts!!

Organizing gapminder into the top 5 values of population in descending order in the year 2007
```{r}
gapminder_top5 <- gapminder %>% 
  filter(year==2007) %>% 
  arrange(desc(pop)) %>% 
  top_n(5, pop)

gapminder_top5
```

Creating the bar chart:
```{r}
ggplot(gapminder_top5) + 
  geom_col(aes(x = country, y = pop))
```

> Q. How do you do this for life expectancy?

```{r}
gapminder_top5_lifeExp<- gapminder %>% 
  filter(year==2007) %>% 
  arrange(desc(lifeExp)) %>% 
  top_n(5, lifeExp)

gapminder_top5_lifeExp

ggplot(gapminder_top5_lifeExp) + 
  geom_col(aes(x = country, y = lifeExp))
```
