---
title: "Class 7: Machine Learning 2"
author: "Derek Zhang (PID: A17819201)"
format: html
toc: true
---

## Principal Component Analysis

## Data Import:

Importing the data below:

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this question?

We can use the dim(df) function. This function returns the dimensions of the data frame in rows x cols.

```{r}
## Complete the following code to find out how many rows and columns are in x?
dim(x)
```

Checking the first 6 rows is done with `head()`

```{r}
head(x)
```

Correcting the row-names problem presented by the row indexes being present as the first row

```{r}
# Note how the minus indexing works
x <- read.csv(url, row.names=1)
head(x)
dim(x)
```

> Q2. The second method of `read.csv(url, row.names=1)` is more robust, as it is a direct read from the source and makes an un-redundant adjustment. When running the other version `x[,-1]`, if ran more than once, there is a risk of deleting more information than desired.
